Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-09-11 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
remane manuscript file and add a footnote

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}
\usepackage{epstopdf}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}	
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}
\usepackage{pbox}
\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}

@@

\institute{Gabriel Riutort-Mayol \at
              Department of Cartographic Engineering, Geodesy, and Photogrammetry, Universitat Polit\`ecnica de Val\`encia, Spain\\
              \email{gabriuma@gmail.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Paul-Christian Bürkner \at
              Department of Computer Science, Aalto University, Finland
           \and
           Michael R.\ Andersen \at
              Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
           \and
           Arno Solin \at
              Department of Computer Science, Aalto University, Finland
           \and
           Aki Vehtari \at
              Department of Computer Science, Aalto University, Finland
}

@@

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{fig12_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{fig11_Posteriors_exI} and \ref{fig12_MSE_exI_inter}, it is seen that the HSGP method yields a good approximation of the exact GP model for both interpolation and extrapolation. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

%In practice, as the high computational demands of exact GPs prevents their use to assess the performance of the approximation, the diagnosis of the approximation as a function of $m$ and $c$ as proposed in Section \ref{subsec_dependency} is recommended.

%The diagnosis consists in comparing the estimated length-scale $\hat{\ell}$ for a given $m$ and $c$, with that minimum length-scale $\ell^*$ that can be accurately approximated, for those given $m$ and $c$, generated by the model of relationships in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig7_lscale_vs_J_vs_c_Matern}. Thus, if $\hat{\ell} < \ell^*$, $m$ needs to increased or $c$ decreased, and this can be iteratively until we get $\hat{\ell} \geq \ell^*$. 
Next, we demonstrate how to apply the diagnostic for approximation quality as a function of $m$ and $c$ as proposed in Section \ref{subsec_dependency}.
Figure \ref{fig13_diagnosis} depicts the diagnosis applied to case study. In the first iteration, we choose $m=20$ and $c=1.5$, which turns out to be an invalid combination of $m$ and $c$, as $c=1.5$ is too small for $m=20$ or $m=20$ is too small for $c=1.5$, as can be seen in the corresponding plot of Figure \ref{fig13_diagnosis}. In the second iteration, $m$ is increased to $m=30$, the HSGP estimates a length-scale of $\hat{\ell}=0.12$, and the minimum length-scale that can be accurately approximated, for those $m$ and $c$, generated by the model of relationships in Figure \ref{fig7_lscale_vs_J_vs_c_Matern} is $\ell^*=0.19$. As $\hat{\ell} < \ell^*$, the approximation is diagnosed to be inaccurate (see Figure \ref{fig12_MSE_exI_inter}-left where the interpolation error of HSGP is higher than that of GP). In the third iteration, $m$ is increased to $m=40$, the HSGP model estimates a length-scale of $\hat{\ell}=0.14$, and the model of relationships suggests a minimum length-scale of $\ell^*=0.15$. As $\hat{\ell}$ is still lower than $\ell^*$, the approximation is diagnosed to be inaccurate, but since $\hat{\ell}$ is quite close to $\ell^*$, the error is expected to be small (see errors in Figure \ref{fig12_MSE_exI_inter}-left). In the fourth iteration, $m$ is increased up $m=60$, the estimated length-scale is $\hat{\ell}=0.14$ and the minimum length-scale generated by the model of relationships is $\ell^*=0.11$, and the approximation is finally diagnosed to be accurate as $\hat{\ell} > \ell^*$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%%!PS-Adobe-3.0 EPSF-3.0
%%%BoundingBox: 19 19 221 221
%%%CreationDate: Mon Sep 29 1997
%%%Creator: programmed by hand (JK)
%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
%\RequirePackage{fix-cm}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}
\usepackage{epstopdf}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}	
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}
\usepackage{pbox}
\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}

@@

\institute{Gabriel Riutort-Mayol \at
              Department of Cartographic Engineering, Geodesy, and Photogrammetry, Universitat Polit\`ecnica de Val\`encia, Spain\\
              \email{gabriuma@gmail.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Paul-Christian Bürkner \at
              Department of Computer Science, University of Stuttgard, Germany \\
              \emph{Most of work was done while at Aalto University}
           \and
           Michael R.\ Andersen \at
              Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
           \and
           Arno Solin \at
              Department of Computer Science, Aalto University, Finland
           \and
           Aki Vehtari \at
              Department of Computer Science, Aalto University, Finland
}

@@

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{fig12_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{fig11_Posteriors_exI} and \ref{fig12_MSE_exI_inter}, it is seen that the HSGP method yield a good approximation of the exact GP model for both interpolation and extrapolation. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

In practice, as the high computational demands of exact GPs prevents their use to assess the performance of the approximation, the diagnosis of the approximation as a function of $m$ and $c$ as proposed in Section \ref{subsec_dependency} is recommended. 
%The diagnosis consists in comparing the estimated length-scale $\hat{\ell}$ for a given $m$ and $c$, with that minimum length-scale $\ell^*$ that can be accurately approximated, for those given $m$ and $c$, generated by the model of relationships in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig7_lscale_vs_J_vs_c_Matern}. Thus, if $\hat{\ell} < \ell^*$, $m$ needs to increased or $c$ decreased, and this can be iteratively until we get $\hat{\ell} \geq \ell^*$. 
Figure \ref{fig13_diagnosis} depicts the diagnosis applied for this case study. In the first iteration, we choose $m=20$ and $c=1.5$, which turns out to be an invalid combination of $m$ and $c$, as $c=1.5$ is too small for $m=20$ or $m=20$ is too small for $c=1.5$, as can be seen in the corresponding plot of Figure \ref{fig13_diagnosis}. In the second iteration, $m$ is increased up to $m=30$, the HSGP estimates a length-scale of $\hat{\ell}=0.12$, and the minimum length-scale that can be accurately approximated, for those $m$ and $c$, generated by the model of relationships in Figure \ref{fig7_lscale_vs_J_vs_c_Matern} is $\ell^*=0.19$. As $\hat{\ell} < \ell^*$, the approximation is diagnosed to be inaccurate (see Figure \ref{fig12_MSE_exI_inter}-left where the interpolation error of HSGP is higher than that of GP). In the third iteration, $m$ is increased up to $m=40$, the HSGP model estimates a length-scale of $\hat{\ell}=0.14$, and the model of relationships brings a minimum approachable length-scale of $\ell^*=0.15$. As $\hat{\ell}$ is still lower than $\ell^*$, the approximation is diagnosed to be inaccurate, but since $\hat{\ell}$ is quite close to $\ell^*$, the error is expected to be small (see errors in Figure \ref{fig12_MSE_exI_inter}-left). In the fourth iteration, $m$ is increased up to $m=60$, the estimated length-scale is $\hat{\ell}=0.14$ and the minimum approachable length-scale generated by the model of relationships is $\ell^*=0.11$, and as $\hat{\ell} > \ell^*$ then approximation is diagnosed to be accurate.
